
---
title: "Introduction to opennaijR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to opennaijR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<div style="height: 45px;"></div>

***opennaijR*** is a reproducible macroeconomic data engineering framework
for Africa. It gives you direct programmatic access to official economic
data from central banks and national statistics bureaus -- starting with
Nigeria and expanding continent-wide. You ask for a dataset by name.
opennaijR connects to the source, cleans the result, and returns a
tidy, analysis-ready object. Every transformation is recorded so your
work is fully reproducible from the moment data enters your workflow.


# Getting Started 

`opennaijR` helps you access Nigerian public data (like inflation, exchange rates, GDP, etc.) in just **three simple steps**:

1. **Load OpennaijR Package**
2. **Discover what datasets are available**
3. **Fetch the dataset you need**

Let’s walk through this step-by-step.


## Step 1: Load the Package

Before doing anything, load the package:

```{r loadPackage, warning=FALSE}
library(opennaijR)
```

That makes all `opennaijR functions` available in your R session.


# Step 2: Discover Available Datasets

Think of `discover_datasets()` as a **catalog of datasets**.

`discover_datasets()` shows you what Nigerian datasets are available and where they come from — so you can decide what to analyze next.
Just like browsing a store before buying something, this function shows you:

* Dataset names
* Dataset IDs
* Data source (e.g., CBN)
* Alternative names (aliases)
* Available variables (columns you can request)

Run:

```{r exploratory, warning=FALSE}
discover_datasets()
```

To see the full list in R console use 'discover_datasets() |> print(n = Inf)`


### What You’ll See

A clean table with columns like:

| dataset_key | dataset_id | source | aliases | variables |
| ----------- | ---------- | ------ | ------- | --------- |


### Why This Is Important

Instead of guessing dataset names, you can:

* See what datasets exist
* Search by keyword
* Check what variables are available


### Search for a Specific Dataset

If you're looking for exchange rate data:

```{r searching, warning=FALSE}
discover_datasets(dataset = "exchange")
```

This filters the catalog to only datasets that match “exchange”.

#### Search Within a Specific Source

If you want to narrow your search to a particular data source (like CBN), you can combine source and dataset.

For example, to find inflation datasets from CBN only:

```{r searching_within, warning=FALSE}
discover_datasets(source = "cbn", dataset = "inflation")
```

This tells opennaijR:

Look only inside the CBN source

Return datasets that match “inflation”

#### Why This Is Useful

Instead of scrolling through everything, you can:

Search by keyword

Limit by source

Quickly discover the exact dataset you need

Think of it like using filters on an online store:

dataset = "inflation" → Search bar

source = "cbn" → Store category filter


# Step 3: Fetch Data

Once you know the dataset name, fetching data is very simple.

To get exchange rates from the Central Bank of Nigeria (CBN):

```{r fetchin_data, warning=FALSE}
exchange_rates <- cbn("exchange_rates")
```

That’s it.

You now have the data stored in:

`exchange_rates`



## What Happens Behind the Scenes?

When you run:

```{r behind_scene, warning=FALSE, eval=FALSE}
cbn("exchange_rates")
```

opennaijR automatically:

1. Builds a data request
2. Checks if you already downloaded it before (cache system)
3. Downloads fresh data if needed
4. Cleans and standardizes the data
5. Saves it for faster reuse next time

So next time you run the same request, it loads instantly.

You don’t need to manage any of this manually.


## Optional: Filter Your Data

You can refine your request.

### Select specific variables:

```{r selection, warning=FALSE}
cbn("exchange_rates", variables = c("currency", "buying_rate"))
```

### Select a date range:

```{r rang, warning=FALSE}
cbn("exchange_rates", from = "2020-01-01", to = "2023-12-31")
```


## Summary

| Task                   | Function                                 |
| ---------------------- | ---------------------------------------- |
| See available datasets | `discover_manifest()`                    |
| Search for a dataset   | `discover_manifest(dataset = "keyword")` |
| Fetch CBN data         | `cbn("dataset_name")`                    |
| Filter by date         | `from`, `to`                             |
| Select columns         | `variables`                              |


## Mental Model

Think of opennaijR like this:

* `discover_datasets()` → **Browse the data catalog**
* `cbn()` "\u2192" **Download the dataset you want**

Simple. Clean. Nigerian data made accessible.



# Filter and Project Data (Clean and Focus)  

After fetching a dataset, it often contains:

* Many rows
* Many columns

But most times, you don’t need everything.

You may want to:

* Filter to specific rows (e.g., only USD)
* Keep only certain columns
* Rename them to something cleaner
* Reorder them for reporting
* Document why you did this

That’s what `apply_projection()` is for.

Think of it as:

> Filtering and reshaping your dataset for your specific task.


## Example: Focus Only on US Dollar Exchange Rates

First, suppose we already fetched exchange rates:

```{r exchange_rates_fetch, eval=FALSE}
exchange_rates <- cbn("exchange_rates")
```

Now we will:

* Keep only **US Dollar**
* Select important columns
* Rename them
* Reorder them
* Record our reason

All in one step:

```{r project, warning=FALSE}
usd_rates <- apply_projection(
  exchange_rates,
  filter = currency == "US DOLLAR",
  cols   = c("date", "currency", "buying_rate", "central_rate", "selling_rate"),
  rename = c(
    buying  = "buying_rate",
    central = "central_rate",
    selling = "selling_rate"
  ),
  order  = c("date", "currency", "buying", "central", "selling"),
  reason = "Focus on USD rates"
)
```


### What Just Happened?

Let’s break it down in simple terms:

| Argument | What It Does                             |
| -------- | ---------------------------------------- |
| `filter` | Keeps only rows matching a condition     |
| `cols`   | Keeps only selected columns              |
| `rename` | Gives columns cleaner names              |
| `order`  | Rearranges column order                  |
| `reason` | Records why you made this transformation |

So instead of technical column names like:

```
ratedate
buying_rate
central_rate
```

You now have clean, readable names:

```
date
buying
central
selling
```



## View the Result

```{r view_rates, warning=FALSE}
head(usd_rates)
```

Now your dataset is:

* Focused (USD only)
* Clean
* Easy to read
* Ready for plotting or reporting

---

### Why `apply_projection()` Is Special

Unlike normal filtering or column selection, this function:

cat("\u2705 Done!\n") Keeps the `opennaijR_tbl` class
cat("\u2705 Done!\n") Tracks what you changed
cat("\u2705 Done!\n") Saves a transformation history

You can inspect the history:

```{r audit_trail, warning=FALSE}
attr(usd_rates, "projection_manifest")
```

This shows:

* When the transformation happened
* What filter was applied
* What columns were kept
* What was renamed
* The reason you provided

That makes your analysis **transparent and reproducible**.


## Derive a New Measure (Create Insight)

After filtering and cleaning your data, the next step is often to create **new variables**.

For example:

* Year-on-year change
* Month-to-month change
* Percentage growth
* Gaps between variables
* Ratios

That’s what `derive_measure()` is for.

Think of it as:

> Creating new economic indicators from existing data.


## Example: Compute Percentage Change in USD Rates

Suppose we already prepared our dataset:

`usd_rates`

Now we want to compute the **percentage change** in exchange rates.

We can do:

```{r}
usd_change <- derive_measure(
  usd_rates,
  pct_change = (buying - lag(buying)) / lag(buying) * 100,
  reason = "Compute daily percentage change in USD buying rate"
)
```

head(usd_rates)

## What Just Happened?

Let’s break it down:

| Component          | What It Does                |
| ------------------ | --------------------------- |
| `pct_change = ...` | Creates a new column        |
| `lag(buying)`      | Uses previous value         |
| `* 100`            | Converts to percentage      |
| `reason`           | Documents why we created it |

So now your dataset contains a new column:

`pct_change`

That tells you how much the USD rate changed compared to the previous period.


## View the Result

```{r}
head(usd_change)
```

You now have:

* Original exchange rate columns
* A new derived column (`pct_change`)
* A recorded derivation history


#### Why `derive_measure()` Is Special

Unlike just writing:

` usd_rates$pct_change <- ...`

`derive_measure()`:

cat("\u2705 Done!\n") Keeps the `opennaijR_tbl` class
cat("\u2705 Done!\n") racks the expression used
cat("\u2705 Done!\n") Records when it was created
cat("\u2705 Done!\n") Stores your reason

You can inspect the derivation history:

```{r}
attr(usd_change, "derive_manifest")
```

This makes your analysis:

* Transparent
* Reproducible
* Auditable

Very important for economic research.


## You Can Derive Multiple Measures at Once

```{r}
usd_extended <- derive_measure(
  usd_rates,
  pct_change = (buying - lag(buying)) / lag(buying) * 100,
  spread = selling - buying,
  reason = "Analyze volatility and bid-ask spread"
)
```

Now you created:

* Percentage change
* Bid-ask spread

In one clean step.

## Mental Model

Your workflow now becomes:
<div class="workflow">
  <code>discover_datasets()</code>
  <span class="arrow">→</span>
  <code>cbn()</code>
  <span class="arrow">→</span>
  <code>apply_projection()</code>
  <span class="arrow">→</span>
  <code>derive_measure()</code>
</div>

